March Madness Basketball Turniment

### 2021-04-15 slides/slides.Rmd
worked on slides

---
title: "March Madness Basketball Turniment"
author: "Sabrina Ball and Joe Despres"
date: "updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 6, echo = FALSE, message = FALSE, warning = FALSE, comment = "", cache = F)

@@

merged <- readr::read_csv(here::here("data", "merged.csv"))
ids <- readr::read_csv(here::here("data", "ids.csv"))
s2021 <- readr::read_csv(here::here("data", "s2021.csv"))
source(here::here("R", "fitting.R"))
```

@@

- Currently, the top 64 teams qualify

@@

- maybe 2022...

@@

- Nobody has ever made a perfect bracket there are $2^{63}$ possible outcomes.

@@

class: inverse, center, middle

# Can we use basketball seasion data to predict this turniment outcomes?

@@

# Predictors

@@

# Distribution of Predictors

```{r fig.width=15, fig.height=10}
hists <- merged %>%
  select(x3fg, fg_percent, rpg, st, to,  bkpg)

histogram <- function(x, t){
  ggplot(data = hists, aes(x)) +
    geom_histogram(bins = 50, color = "black", fill = "gray80") +
    theme_light(base_size = 20) +
    labs(title = paste0("", t), x = "", y = "") +
    NULL

@@

@@

# Poisson Model

```{r}
equatiomatic::extract_eq(model = fit_poisson,
                         use_coefs = FALSE,
                         show_distribution = TRUE,
                         wrap = TRUE,
                         terms_per_line = 2,
                         intercept = "beta")
```

@@

# Multinomeal Model

---
class: center 
### Output and Interpretation

```{r}
table_output <- function(obj, rows){
  broom::tidy(obj) %>% 
  mutate(p.value = as.character(p.value)) %>%
  mutate_if(is.numeric, round, 3) %>% 
  mutate(p.value = as.numeric(p.value))  %>% 
  kable(col.names =  c("Term", "Estimate", "Std. Error", "T-Stat", "P-Value")) %>% 
  kable_styling(font_size = 15) %>% 
  row_spec(rows, color = "white", background = "#025cba")
}

---
class: center 
## Output and Interpretation

---
class: center 
## Output and Interpretation

---
class: center 
## Output and Interpretation

---
class: center 
## Output and Interpretation

---
class: center 
## Output and Interpretation

---
class: center 
## Output and Interpretation

---
class: center
# Scoring

$$\LARGE\sum_{i=1}^{63} (p_i * \mathds{1}_{correct prediction}  -p_i * \mathds{1}_{incorrect perdiction}) = score$$

@@

# Performance

@@

@@

>>>>

---
title: "March Madness Basketball Turniment"
author: "Sabrina Ball and Joe Despres"
date: "updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=15,
                      fig.height=12, 
                      echo = FALSE, 
                      message = FALSE,
                      warning = FALSE, 
                      comment = "")

@@

merged <- readr::read_csv(here::here("data", "merged.csv"))
ids <- readr::read_csv(here::here("data", "ids.csv"))
s2021 <- readr::read_csv(here::here("data", "s2021.csv"))
source(here::here("R", "fitting.R"))
theme_set( theme_light(base_size = 25))
```

@@

- Now, its restricted to the top 64 NCAA Division I Teams

@@

@@

- Nobody has ever made a perfect bracket and there are $2^{63}$ possible outcomes.

@@

class: inverse

# Using basketball seasion data to predict this year's turniment outcomes requires us to:

### Design predive models to make out of sample predictions

### Derrive a scoring metric

### Evaluate and tune the model

### We made several models, In the interstst of time we are going to discuss the logistic because its the most straight-forward

```{r fig.width=15, fig.height=12}
hists <- merged %>%
  select(x3fg, fg_percent, rpg, st, to,  bkpg)

histogram <- function(x, t){
  ggplot(data = hists, aes(x)) +
    geom_histogram(bins = 55, color = "white", fill = "black") +
    theme_light(base_size = 25) +
    labs(title = paste0("", t), x = "", y = "") +
    NULL

@@

# Real Quick Descriptive Statisticcs

@@

- Normal-ish distributions

- No catastrophic skews or heavy tails

- Ideally we would have had steaks and turn-overs per game rather than season total. Not all played the same game

@@

Summary

```{r}
row_name <- c("Intercept",
  "Three Pointers", "O* Three Pointers", 
          "Field Goal Percentage", "O* Field Goal Percentage",
          "Free-throw percentage", "O* Free-throw percentage",
          "Rebounds", "O* Rebounds",
          "Steals", "O* Steals",
          "Turnovers", "O* Turnovers",
          "Blocks", "O* Blocks"
          )

cols <- c("Term", "Estimate", "Std. Error", "T-Stat", "P-Value")

table_output <- function(obj, rows){
  broom::tidy(obj) %>%
  mutate(p.value = as.character(p.value), term = row_name) %>%
  mutate_if(is.numeric, round, 3) %>%
  mutate(p.value = as.numeric(p.value))  %>%
  kable(col.names = cols) %>%
  kable_styling(font_size = 15) %>%
  row_spec(rows, color = "white", background = "#025cba") %>% 
  add_footnote("Where O* is the Mertric associated with their opposing team", notation = "none")
}
  broom::tidy(logistic_fit) %>%
  mutate(p.value = as.character(p.value), term = row_name) %>%
  mutate_if(is.numeric, round, 3) %>%
  mutate(p.value = as.numeric(p.value))  %>%
  kable(col.names = cols) %>%
  kable_styling(font_size = 15) %>%
  # row_spec(rows, color = "white", background = "#025cba") %>% 
  add_footnote("Where O* is the Mertric associated with their opposing team", 
               notation = "none")

```

---
class: center
Summary

```{r}

---
class: center
Summary

---
class: center
Summary

---
class: center
Summary

---
class: center
Summary

---
class: center
Summary

---
class: center
Summary

---
## Scoring

$$\sum_{i=1}^{n} [\mathbf {1} _{\{correct\ prediction\}} \hat{p_i} -\mathbf {1} _{incorrect\ perdiction}\hat{p_i} ] = Model\ Score$$

- Basketball games are often quite close and odds in betting markets are often close to 1:1.

- Our goal Goal is to weight correct prediction higher when it assigns a higher probability. Conversely a lower score to an incorrect prediction.

- $E(Model\ Score) = 0$ if our predictions are no better than randomly guessing.

@@

## Scoring

$$\sum_{i=1}^{n} [\mathbf {1} _{\{correct\ prediction\}} \hat{p_i} -\mathbf {1} _{incorrect\ perdiction}\hat{p_i} ] = Model\ Score$$

Minimal Example

```{r}
col_fix <- function(x){
  x <- str_replace(x, "_", " ")
  x <- str_to_title(x)
  x
}

logistic_model_score <- readRDS(here::here("cache", "logistic_model_score.RDS")) %>% 
  drop_na() %>% 
  head(10) %>% 
  mutate_if(is.character, str_to_title) %>% 
  mutate_if(is.numeric, round, 2) %>% 
  mutate(phat = if_else(predicted_winner == winner, 
                          predicted_winner_probs, predicted_loser_probs)) %>% 
  rename_all(col_fix) %>%
  rename( `$\\hat{p}$` = Phat) %>% 
  select(Game, `Predicted Winner`, `Winner`, `$\\hat{p}$`, Score)
```

```{r}
logistic_model_score %>% 
  kable(align = "r") %>% 
  kable_styling(font_size = 15) 
```

@@

## Scoring

$$\sum_{i=1}^{n} [\mathbf {1} _{\{correct\ prediction\}} \hat{p_i} -\mathbf {1} _{incorrect\ perdiction}\hat{p_i} ] = Model\ Score$$

Correct Predictions

```{r}
logistic_model_score %>% 
  kable(align = "r") %>% 
  kable_styling(font_size = 15) %>%
  column_spec(c(1:5), color =  ifelse(logistic_model_score$Score > 0, "white", "black"), 
              background = ifelse(logistic_model_score$Score > 0, "#025cba", "#f0f0f0"))
```

@@

## Scoring

$$\sum_{i=1}^{n} [\mathbf {1} _{\{correct\ prediction\}} \hat{p_i} -\mathbf {1} _{incorrect\ perdiction}\hat{p_i} ] = Model\ Score$$

Score

```{r}
logistic_model_score %>% 
  kable(align = "r") %>% 
  kable_styling(font_size = 15) %>% 
  column_spec(c(5), color =  ifelse(logistic_model_score$Score > 0, "white", "black"), 
              background = ifelse(logistic_model_score$Score > 0, "#025cba", "#ff3700"))
```

```{r fig.align='right'}
paste0("Model Score = ", sum(as.numeric(logistic_model_score$Score))) %>% 
  kable(col.names = "$\\sum_{i}Score$") %>% 
  kable_styling(font_size = 15) %>% 
  row_spec(1, color = "white", background = "#025cba")
```

@@

```{r}
perform <- readRDS(here::here("cache", "logistic_model_score.RDS")) %>% 
  mutate(outcome = if_else(score > 0, "Correct", "Incorrect")) %>% 
  drop_na() 
```

```{r fig.height=10}

a <- perform %>%
  ggplot(aes(x = abs(score), fill = outcome)) + 
   geom_histogram(alpha=0.7, position="identity", bins = 10) + 
  labs(title = "Histogram of Scores", x = "Absolute Value of Score") +
  scale_fill_manual(values = c("#025cba", "#ff3700"))

b <- perform %>% 
  ggplot(aes(as.factor(outcome), fill = outcome)) + 
  geom_bar(alpha = 0.7) + 
  labs(title = "", x = "") + 
  scale_fill_manual(values = c("#025cba", "#ff3700")) +
  geom_text(size=10, color = "grey80", stat = 'count', aes(label =..count.., hjust = 1.5)) + 
  coord_flip()

a/b
```

```{r}
t2 <- perform %>% 
  summarise(totalscore = sum(score),
            se = sd(score), 
            t = totalscore/se,
              p = pt(t, nrow(perform)-1, lower.tail = F)) %>% 
  t()

score <- perform %>% 
  group_by(outcome) %>% 
  summarise(totalscore = sum(score)) %>% 
  mutate(totalscore = round(totalscore, 2))

t1 <- rbind(score, c("Total", sum(score$totalscore))) 
```

```{r}
knitr::kable(list(t1, t2)) %>% 
  kable_styling(font_size = 10)

```

---
class: inverse

@@

### Logistic Regression performs well. Given measured statists.

### However out model is independent of the pervious game which would aplify shortcommings.
